Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-15 model.tex
rewrite of the first subsection of section 4

\section{Modeling }\label{sec:model}
{\color{blue} EDITABLE}

Section outline:

\begin{enumerate}
\item \cmark section intro
\item \cmark fix order from simplest to most complex
\item Description of DCE and parameter estimation  
\item\cmark Description of auto ARIMA 
\item \cmark Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
\item\cmark Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.

\end{enumerate}

In this section we describe the four different forecasting schema used in this study. In Section \ref{sec:simple} we outline two primitive strategies which often out perform sophisticated prediction strategies in the case of highly complex signals. In Section \ref{sec:arima} we describe Autoregressive integrated moving-average (ARIMA) a commonly used linear forecasting scheme. In Section \ref{sec:lma} we describe a simple nonlinear deterministic forecasting scheme, Lorenz Method of Analogous (LMA) for capturing and exploiting nonlinear dynamics present in a time series. And finally in Section \ref{sec:accuracy} we discuss the error metric we use to evaluate each models predictive accuracy.

\subsection{Primitive Strategies to combat complexity: Random Walk and \naive {\color{blue} EDITABLE}}\label{sec:simple}
In this section we cover two very simple forecast models: Random Walk and \naive. With a Random Walk predictor to forecast at time $i$ we simply use the last observed measurement: $$p_i = X_{i-1,obs}.$$ The \naive  prediction strategy is to forecast the rolling average up to that point. i.e., the prediction $p_i$ is simply the average of all observed points up to that point: $$p_i = \sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}.$$

While both of these methods are simplistic they are not without merit. When time series possess very little predictive structure,i.e., they exist on the very top of the complexity spectrum these two methods can sometimes be \emph{on-average} very good choices for a forecasting method, as we will see in our results. An example is forecasting currency exchange rates. In \cite{rwMeese,rwCCE} the authors show that various econometric techniques and prediction models fail to consistently outperform Random Walk predictions in forecasting various exchange rates. This is the case because these signals are volatile, noisy and possess very little information structure but they don't \emph{on-average} vary a great deal. That is the day-to-day exchange rates \emph{on-average} don't vary very much so guessing the last known exchange rate is often a very good approximation to the new exchange rate. [[Not a huge fan of this wording.]] The Na\"ive forecasting scheme can be a great prediction strategy if the time series has very little forward information structure, adheres to a unimodal distribution and has very little variance. In this circumstance, forecasting the mean will on average get you in the ball park of the forecast without taking into account the lack of structure. The \naive~ forecasting scheme can be thought of as a filtering mechanism which compresses a signal onto a single point, effectively filtering out both complexity and structure. However, if very little predictive structure is present in a signal then this filtering mechanism effectively mitigates the complexity.

While the simplicity of these methods is advantageous on average when a time series exists on the high side of the complexity spectrum, their simplicity can also be a disadvantage and cause the methods to be inconsistent in their forecasting ability as a consequence of not being dynamic or taking into account the distribution or structure of the time series. Both of these methods fail spectacularly in two specific circumstances that lie on the fringes of their strengths. Fortunately, they somewhat compliment each other, i.e., the weakness of Random Walk and \naive~ are covered by the other primitive method.

The situation in which Random Walk fails is when a time series changes drastically on a regular basis. The simplest example of this would be a period 2 time series which alternated between 0 and 1 at each time step. At every time step Random Walk would precisely predict out of phase, and as a result would accrue an error of 1 at every time step ($|p_i-X_{i,obs}|=1$. In comparison, the \naive method would forecast 0.5 at each time step and would thus accrue an error of 0.5 at every step. So on-average \naive would do twice as well as Random Walk, and as you could imagine a dynamic method would captured this periodicity  and exploited this predictive structure could do much much better, such a method is introduced in Section \ref{sec:lma}.

The \naive forecast fails when you have a signal which shifts between two or more regimes on a regular basis and the regimes are separated by a significant amount. In this case the \naive forecast will predict between the regimes at every time step eventually accruing massive error. To see how Random Walk can cover this weakness imagine a time series which switches between more than one regime but stays in each regime for several time steps, e.g., a regularly sampled noisy square wave, this behavior can be seen in some portions of the \svd trace seen in Figure \ref{fig:svd}. Random walk in this case could react to the regime shift within one time step, switch to the new Regime and begin forecasting nicely, where as \naive would just continue to predict between the modes of the multimodal distribution resulting in poor forecasting accuracy all of the time.

So clearly these primitive methods have their downfalls, but it is important to keep in mind that when a time series exists on the top of the complexity spectrum they have little predictive structure to exploit and often a primitive method like Random Walk or \naive~ will consistently out perform methods which rely on predictive structure. This is explored further in Section \ref{sec:accuracy} and \ref{sec:results}

>>>>

\section{Modeling }\label{sec:model}
% 
% {\color{blue} EDITABLE}
% 
% Section outline:
% 
% \begin{enumerate}
% \item \cmark section intro
% \item \cmark fix order from simplest to most complex
% \item Description of DCE and parameter estimation  
% \item\cmark Description of auto ARIMA 
% \item \cmark Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
% \item\cmark Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results. 
% 
% \end{enumerate}

In this section, we describe the four different forecasting schema
used in this study, as well as the error metric used to evaluate their
predictive accuracy:
\begin{itemize}
\item The \emph{random-walk} method, which uses the previous value in
  the observed signal as the forecast,

\item The \emph{naive} method, which uses the mean of the
  observed signal as the forecast,

\item The \emph{ARIMA} method, a common linear forecast strategy, and

\item The \emph{LMA} method, which uses a near(est) neighbor forecast
  strategy on a dynamical reconstruction of the signal.
\end{itemize}
ARIMA is based on standard autoregressive techniques.  LMA is designed
to capture and exploit the deterministic structure of a signal from a
nonlinear dynamical system.  The first two methods, somewhat
surprisingly, often outperform sophisticated prediction strategies in
the case of highly complex signals, as discussed below.

\subsection{Two Simple Prediction Strategies}
\label{sec:simple}

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the value $p(i)$ at time $i$ is predicted to
be $$p_i = X_{i-1,obs}$$ The \naive ~prediction strategy uses a
rolling average to generate the forecast: the prediction $p_i$ is
simply the average of [[all]] [[no window?  if so, then it's not a
    ``rolling'' average]] observed points up to that point: $$p_i =
\sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}$$

While both of these methods are simplistic, they are not without
merit.  For a time series near the high end of the complexity
spectrum---i.e., one that possesses very little predictive
structure---these two methods can actually be the best choice.  In
forecasting currency exchange rates, for instance, sophisticated
econometric techniques and prediction models fail to consistently
outperform the random-walk method~\cite{rwMeese,rwCCE}.  These signals
are volatile, noisy, and possess very little predictive structure, but
they do not---on the average---vary a great deal, so guessing the last
known value is not a bad strategy.  If a signal has a unimodal
distribution with low variance, the \naive ~prediction strategy will
perform quite well---even if the signal is highly complex---simply
because the mean is a good approximation of the future behavior.
Also, a temporal average effects a low-pass filtering operation.  In a
signal with very little predictive structure, then, the \naive
~prediction strategy effectively mitigates the complexity.

Both of these methods have significant weaknesses, however, because
they do not model the temporal patterns of the time series or even the
distribution of its values.  Thus, they cannot track changes in that
structure, which causes them to fail in a number of important
situations.

Random-walk strategies are a particularly bad choice for time series
that change drastically at every time step.  In the worst case---a
large-amplitude square wave whose period is equivalent to twice the
sample time---a random-walk prediction would be exactly 180 degrees
out of phase with the true continuation.  The \naive ~method would be
better in this situation, since it would always predict the mean.  It
would, however, fail when a signal has a number of long-lived regimes
that have significantly different means.  In this situation, the
inertia of the \naive ~method's rolling mean is a liability and the
random-walk method's nimbleness is an advantage, allowing it to
react quickly to regime shifts.

Of course, a method that could capture and exploit periodic patterns
would be far more effective in these examples.  The LMA method
introduced in Section~\ref{sec:lma} is designed to do exactly that.
[[Couldn't one think about ARIMA as moving in that direction, too?]]
Note, though, that periodic patterns appear only in signals that are
at the low end of the complexity spectrum.  If a signal contains
little predictive structure, forecast strategies like LMA [[and
    ARIMA?]] have nothing to exploit and thus will be beaten out by
the two simple strategies described in this section.  This is explored
further in Sections~\ref{sec:accuracy} and~\ref{sec:results}
